\chapter{INTRODUCTION}

% \pagestyle{plain}

% no \PARstart
In recent years, wireless sensor networks (WSN)
have become increasingly popular in playing the
role of real world modalities. A WSN consists of a
large number of tiny sensor nodes capable of
interfacing with the real world, communicating
wirelessly and performing limited computation.
These tiny sensor nodes are low cost, low power and
are easily deployable and can be readily networked
and hence offers opportunity for a wide range of
applications. Most popular areas for WSN
applications are health, military, environment,
home and security.
%
\section{Scope and Challenges}
\label{scope-challenges}
Due to tight integration with the physical world,
sensor networks present significant challenges to
the application development. As a result, most of
the WSN applications till date focus on simple
data-gathering and control designed to work on
specific network and hence resulting in monolithic
applications strongly coupled with protocols and
network used. But as sensor network applications
become pervasive, such monolithic and ad hoc
approaches does not work and a systematic
application design method based on standards and
higher level abstractions is required [22]. To
simplify sensor network application development,
the need for programming abstractions such as a
middleware framework is well acknowledged. WSN
middleware can be considered as a software
infrastructure that resides above operating system
and below the application abstracting low-level
functionality by providing high-level abstractions.
A complete middleware solution should provide a
holistic view of the network, services for data
collection, aggregation and control while
efficiently and adaptively managing system
resources as per application's requirements. \\ 
%%
%% Resource management (DIRL+COIN)
WSN nodes are remarkably constrained in terms of
their resources viz. energy, computational power
and radio bandwidth. WSNs normally operate in
uncertain and dynamic environments where the state
of the system changes considerably over time.
For example, in data collection applications, uncertainty exists due to
intermittent links or traffic conditions. Moreover,
the network itself is dynamic due to such events as
node mobility and depleted battery. WSN
applications need to cope with such dynamicity and
uncertainty inherent in sensor networks, while
simultaneously trying to achieve application's
requirements for QoS and optimization goal.
Consequently, adaptive resource management is a key
to any successful middleware
solution enabling such applications.\\ 
Resource management includes initial
sensor-selection and task allocation as well as
runtime adaptation of allocated task/resources.
There are many proposed middleware solutions that
have advocated strong need for proactive adaptation
of resources [5,14,28], but there are only few that
have actually tried to resolve the issue of
enabling adaptive resource management for WSN
applications. This problem of resource
management/adaptation can be described as follows:
\\ \emph{Given application structure, QoS
requirements and current system state, what is the
best way of allocating tasks to resources so that 
given system-wide, application-driven, global
parameters can be optimized.}\\ In the above, the
application structure is in the form of underlying
tasks and their interactions. QoS requirements
include such constraints as latency, reliability,
coverage etc. while the current state of the system
is defined by parameters like mobility, energy
availability, and neighboring nodes. The parameters
to be optimized include energy, bandwidth, and
network lifetime. Without loss of generality, we
illustrate the resource management problem using a
simplified object/entity tracking application.
Object tracking application can be considered to
consist of following tasks: 1) sample- sense the
environment (e.g. signal strength of a moving
object). 2) transmit (Tx)- transmit a message to
next hop towards the base-stations. 3) receive
(Rx)- turn radio to receive mode to listen for
incoming messages. 4) aggregate- aggregate two or
more local and remote same target readings into
single reading (e.g. data triangulation for better
position estimation or mapping to a known track or
simply \emph{last value} aggregation function). 5)
sleep- put CPU and radio in sleep mode to minimize
battery consumption. State representation may
consist of the following variables: have one or
more neighbors, successful in recent sampling,
successful in recent receive, signal strength (or
quality of reading). QoS requirements here may
include quality of signal, tracking coverage area
as well as maximum allowed latency. Our goal in
this case is to optimize energy usage among all
sensor nodes. Thus the goal of our resource
management framework is to schedule and allocate
tasks on each sensor node in the system, so that
energy usage among all sensor nodes is minimized
while fulfilling the coverage and latency
requirements of the application.\\
%%
%% Sparse WSN
Most sensor network applications studied today consists of a large number of sensor nodes 
deployed over a geographical area. Sensors use
multi-hop communication to send data acquired from
the external environment to a sink node or to an
Access Point (AP) in the infrastructure. However,
several applications do not require fine-grained sensing. Examples of such
applications include monitoring of weather
conditions in large areas, air quality in urban
scenarios, terrain conditions for agriculture, and
so on. In this case, it is possible to consider a
\emph{sparse wireless sensor network}, i.e., a WSN
where the density of nodes is so low that they
cannot communicate with each other through
multi-hop paths, or even directly. In order to make
communication feasible, data collection in sparse
WSNs can be accomplished by means of \emph{mobile
data collectors} (MDCs). MDCs are special mobile
nodes responsible for data gathering and/or
dissemination. They are assumed to be powerful in
terms of data storage and processing capabilities,
and are not energy constrained. However, the data
collection paradigm in sparse WSNs with MDCs is
different, and introduces significant challenges
including contact detection and energy
conservation. \\
Communication between an MDC and each sensor node
takes place in two phases. First, sensor node
discovers the presence of the MDC in its
communication range. Then, it transfers collected
data to the MDC while satisfying any required
reliability constraints. Unlike MDCs, sensor nodes
have a limited energy budget, so that the
data-collection process has to be energy efficient
in order to prolong their network lifetime. In
addition, such energy-conserving mechanisms should
not compromise the timeliness of communication.
This is critical especially when the MDC has only a
short contact time with sensors, and also in the
case when such contacts cannot be predicted
accurately. In fact, a major problem in data
collection is that sensor nodes usually do not have
\emph{a priori} knowledge of the MDC mobility
pattern. Furthermore, even in cases where the
arrivals can be predicted, there is a chance that
the MDC contacts can be affected by delays or can
change their rate. Hence robust and flexible
mechanisms have to be defined in order to adapt to
operating conditions autonomously.\\
%
% % Communication Paradigm
Support for a scalable and robust communication
paradigm is another important characteristic of a
middleware solution. As all sensor network
applications are data-driven, middleware should
ideally support a data-centric model for task and
data dissemination. Middleware should allow
application specific in-network processing and data
filtering to enable application assisted routing,
aggregation, data fusion and collaborative
information processing. %
\section{Contributions}
\label{contributions}\subsection{Adaptive Resource Management Using Distributed Independent Reinforcement Learning}\label{intro:dirl}
\subsection{Ensuring Global Optimization With Multi-Tier Reinforcement Learning}\label{intro:coin}\subsection{Resource-aware Data Collection in Sparse Wireless Sensor Networks}\label{intro:dirl-sparse}

\subsection{Design of Adaptive Middleware Using Directed Diffusion}\label{intro:diffusion}
%
\section{Organization of Dissertation}\label{organization}
